---
output:
  html_document: default
  pdf_document: default
  word_document: default
---
```{r echo = FALSE, results ='asis', message=FALSE, warning=FALSE}
library(jsonlite)
jsonFile = "BCO-DEA.json"
bco <- fromJSON( jsonFile)
doc_title <- cat(" BCO Example",bco$provenance_domain$name)
doc_author <- 'Paul Walsh'
doc_date <- Sys.Date()
```

---
title: `r doc_title`
author: `r doc_author`
date: `r doc_date`
output: html_document
---
# Bio Compute Objects DEA Example
## Provenance
```{r echo = FALSE, results ='asis'}
cat('* Version:',bco$provenance_domain$version)
cat('\n* Created:',bco$provenance_domain$created)
cat('\n* Modified:',bco$provenance_domain$modified)
cat('\n* Keywords:',paste( unlist(bco$description_domain$keywords$value), collapse=', '))
cat('\n* Current Status:',bco$provenance_domain$review$status[1])
cat('\n* Licence:',bco$provenance_domain$license$address)
```

## Contributers
```{r echo = FALSE, results ='asis'}
cat(paste(bco$provenance_domain$contributors$entity$contribution, bco$provenance_domain$contributors$entity$name,  bco$provenance_domain$contributors$entity$affiliation, bco$provenance_domain$contributors$entity$email, sep=', ', collapse='<br>'))
```

## Review
```{r echo = FALSE, results ='asis'}
cat(paste(unlist(bco$provenance_domain$review$comment), bco$provenance_domain$review$reviewer$entity$name,  bco$provenance_domain$review$reviewer$entity$affiliation, bco$provenance_domain$review$reviewer$entity$email, sep=', ', collapse='<br>'))
```

## Extension
```{r echo = FALSE, results ='asis'}
cat('GitHub Repo:',bco$extension_domain$github_extension$github_repository, "<br>")
cat('GitHub link:',bco$extension_domain$github_extension$github_URL)
```

### FHIR
```{r echo = FALSE, results ='asis'}
cat(paste(unlist(bco$extension_domain$FHIR_extension$FHIRendpoint_Ids), bco$extension_domain$FHIR_extension$FHIRendpoint_Resource,  bco$extension_domain$FHIR_extension$FHIRendpoint_URL, sep=', ', collapse='<br>'))
```

## Analysis Pipeline Steps
```{r echo = FALSE, results ='asis'}
cat(paste(unlist(bco$description_domain$pipeline_steps$tool$step_number), bco$description_domain$pipeline_steps$tool$name,  bco$description_domain$pipeline_steps$tool$description, bco$description_domain$pipeline_steps$tool$version, sep=', ', collapse='<br>'))
```

### Required Libraries
This section aids reproducibility by listing required libraries and ensuring that they are loaded. It loads the BioCondoctor package listed in the BCO:
```{r echo = FALSE, results ='asis'}
cat(paste(unlist(bco$execution_domain$software_prerequisites$name), bco$execution_domain$software_prerequisites$version, sep=', ', collapse='<br>'))
```

#### Installation
```{r echo = FALSE, results='hide', message=FALSE, warning=FALSE}
# Import package from BCO
fileName= bco$execution_domain$software_prerequisites$uri$address[1]
source(fileName)
#biocLite("edgeR", type="binary")
#biocLite("ggplot2")
#biocLite('statmod')
#biocLite('pheatmap')
library("edgeR")
library("ggplot2")
library("pheatmap")
library("ggfortify")
```

```{r echo = FALSE, message=FALSE, warning=FALSE, results ='asis'}

cat("Packages:", paste((.packages()), sep= "," ))
cat("<br>EdgeR Version:", paste(packageVersion("edgeR") ))
cat("<br>ggplot2 Version:", paste(packageVersion("ggplot2") ))
cat("<br>pheatmap Version:", paste(packageVersion("pheatmap") ))
cat("<br>ggfortify Version:", paste(packageVersion("ggfortify") ))
```

### Reading input files
Read files specified in BCO.

#### Reading the clinical data
Get clinical input file source from BCO:
```{r echo = FALSE, results='asis', message=FALSE, warning=FALSE}
fileName= bco$io_domain$input_subdomain$subject$URI$address[2]
cat(fileName, "<br>")
clinicalTable = read.table(fileName,header=TRUE,sep = '\t',row.names = 1)
#clinicalTable = read.table('phenotypical_final.txt',header=TRUE,sep = '\t',row.names = 1)
# verifying if the table was correctly read. The function allows to see the first 6 rows.
#head(clinicalTable)
# Checking the variables
#summary(clinicalTable)
```

                           
#### Reading the gene expression data
Get gene expression data input file source from BCO:
```{r echo = FALSE, results='asis', message=FALSE, warning=FALSE}
fileName= bco$io_domain$input_subdomain$subject$URI$address[1]
cat(fileName, "<br>")
exp = read.table(fileName,header=TRUE,sep = '\t',row.names = 1)
#exp = read.table('countable_final.txt',header=TRUE,sep = '\t',row.names = 1)

#Checking the size of the table
#dim(exp)

#Checking the first rows
#head(exp)
#summary(exp)
```

## Modeling the data
Pull the model from the BCO:
```{r echo = FALSE, results='asis', message=FALSE, warning=FALSE}
# Two variables were included in the statistical model:
# 1) the variable of interest - clark_level
# 2) a source  of variation - tissue site
variables= bco$parametric_domain$model$factors
cat(variables, "<br>")
design <- model.matrix(as.formula(paste("~",variables)), data = clinicalTable)
```

## Processing the count table
Get parameters such as counts per million from the BCO.
```{r echo = FALSE, results='asis', message=FALSE, warning=FALSE}
# Creating an object that combines the expression data (count table, named 'exp') and clinical data. Required for analysing with edgeR.
y=DGEList(counts = exp, samples = clinicalTable)
cpmMin = variables= bco$parametric_domain$cpm$min
cat("<br>Counts per million:", cpmMin, "<br>")
# Removing sequences with very low frequency
keep <- names(which(rowSums(cpm(y)) >= cpmMin))
y <- y[keep, , keep.lib.sizes=FALSE]

# Normalization
y <-  (y)

# Estimating dispersion
y<- estimateGLMRobustDisp(y,design)
```

### Observing the processed data
```{r results='hide', message=FALSE, warning=FALSE}
# Plot Biological Coefficient of Variation
# "Plot the genewise biological coefficient of variation (BCV) against gene abundance (in log2 counts per million)."
plotBCV(y)

# Plotting Multidimensional scaling plot of distances between gene expression profiles
# "Plot samples on a two-dimensional scatterplot so that distances on the plot approximate the typical log2 fold changes between the samples."
plotMDS(y, col=c('green','purple')[clinicalTable$clark_level], pch=c(21,23,22)[clinicalTable$tumor_tissue_site])
legend('topright', col=c('green','purple'), legend=levels(clinicalTable$clark_level), pch = 16, cex = 0.7)
legend('bottomright', legend=levels(clinicalTable$tumor_tissue_site), pch = c(21,23,22), cex = 0.7)

```

## Fitting the model to the processed read count data and conducting statistical tests
```{r results='hide', message=FALSE, warning=FALSE}
fit <- glmFit(y, design)
lrt <- glmLRT(fit, coef=2)

#Visualising the sequences differently expressed between levels III and IV
plotMD(lrt, main = 'Melanoma Clark Level')
abline(h=c(-1, 1), col="grey")
```

```{r}
res=lrt$table
res$FDR=p.adjust(res$PValue, method="BH")
res=res[order(res$FDR),]

#displaying the top 25 results
res[1:25,]

```
## Heatmap
Get FDR from BCO.
```{r}
logcpm <- cpm(y, prior.count=2, log=TRUE)
fdrValue = variables= bco$parametric_domain$fdr$value
# Filtering the sequences differently expressed
select=rownames(res)[which(res$FDR<=fdrValue)]

# Applying Euclidean distance to cluster samaples and the 10 genes differently expressed
pheatmap(logcpm[select,], cluster_rows=T, show_rownames=T,
         cluster_cols=T, annotation_col=clinicalTable[,c( "clark_level","tumor_tissue_site")], 
         main="Sequences differently expressed")

# It is possible to observe that the gene expression can separate the majority of samples by Clark Level
```
